Understanding meta-analysis through data simulation with applications to power analysis

### 2023-09-06 documents/paper/paper.tex
updating after revision

@@

\abstract{%
Meta-analysis is a powerful tool to combine evidence from existing literature. Despite several introductory and advanced materials about organizing, conducting, and reporting a meta-analysis, to our knowledge, there are no introductive materials about simulating the most common meta-analysis models. Data simulation is essential for developing and validating new statistical models and procedures. Furthermore, data simulation is a powerful educational tool for understanding a statistical method. In this tutorial, we show how to simulate \rev{equal-effects}, random-effects, and meta-regression models and illustrate how to estimate statistical power. In the supplementary materials, we extended these simulations also for multilevel and multivariate models. All materials associated with this article can be accessed on Open Science Framework (\url{https://osf.io/54djn/}).
}

@@

A meta-analysis is an essential tool for combining knowledge from multiple studies quantitatively. Meta-analysis is commonly used together with a systematic review of the literature. The meta-analysis has several advantages. Firstly, it allows combining evidence from multiple studies assigning more weight to studies with lower estimation variability. \color{red} Then using meta-regression, it is possible to include variables (i.e., moderators) to explain the observed heterogeneity (Borenstein, Hedges, Higgins, \& Rothstein, 2009, pp. 187--203). More recently, location-scale models have been developed to include predictors also on the residual heterogeneity (i.e., Viechtbauer \& López-López, 2022). \color{black} Finally, considering the replication crisis, there are statistical methods to determine the presence and extent of the publication bias. Despite the advantages, meta-analysis implementation is not always straightforward, especially for complex data structures. Additionally, the fact that there are several introductory and advanced resources to understand meta-analysis (Borenstein et al., 2009; Harrer, Cuijpers, Furukawa, \& Ebert, 2021; Schmid, Stijnen, \& White, 2022), to our knowledge, there are no introductory resources about how to simulate realistic meta-analytic data. Simulating data has several advantages because it requires understanding the statistical method and the data-generation process. Furthermore, data simulation is the primary tool when it comes to evaluating a new analysis method, estimating the statistical power, or understanding the long-run behavior of our data generation process (Gelman \& Hill, 2006, pp. 155--176; Gelman, Hill, \& Vehtari, 2020, pp. 69--76; Ingalls, 2011). A recent paper by DeBruine and Barr (2021), which deeply inspired the current work, proposed a stimulating way to understand linear \textcolor{red}{mixed-effects models} via data simulation. Simulating data is also a powerful educational tool within this framework. For these reasons, this work aims to introduce the basic concepts of meta-analysis and Monte Carlo simulations for \textcolor{red}{equal-effects}, random-effects, and meta-regression models with applications also to statistical power calculation. In the first section, we will introduce basic concepts of the meta-analysis that are useful for setting up the simulation. We will evaluate the effect size, variance calculation, and the \textcolor{red}{equal} vs.~random effects model distinction. Then we will describe how to simulate data for these models and simulate a meta-regression with categorical and numerical predictors. Finally, we will introduce the power analysis extending the previous examples to estimate the statistical power. We used the R statistical programming language (version 4.2.3; R Core Team, 2023). We assume the reader is familiar with the basic concepts of R, but core functions will be explained. Code and materials are available on the OSF repository (\url{https://osf.io/54djn/}). \color{red} A theoretical introduction, simulation examples for multivariate and multilevel models, and more details about the coding approach are available in the supplementary materials. \color{black}

@@

\color{red} As an example that will be used throughout the paper, we consider the efficacy of memory training in improving memory performance during a cognitive task. The typical primary study will collect data from a group of participants receiving the memory training (\emph{experimental group}) and another group receiving a control treatment (\emph{control group}). The focus of the meta-analysis is collecting multiple studies with similar aims and methods and estimating the average effect of memory training. Despite differences in the type of cognitive task or experimental setup, each primary study collects an \emph{experimental group} (\(n_{T}\)) and \emph{control group} (\(n_{C}\)) and computes the average performance (\emph{experimental group} \(\overline T\) and \emph{control group} \(\overline C\)) and standard deviations (\emph{experimental group} \(s_{T}\) and \emph{control group}
\(s_{C}\)). \color{black}

@@

The first step of a meta-analysis is to extract information from included studies. This common measure should give an immediate idea of the direction (i.e., the treatment improves or reduces performance) and the size of the effect. Standardized effect size measures (see Lakens, 2013 for an overview), such as the Standardized Mean Difference (SMD), which could be estimated by Cohen's \(d\) (Cohen, 1988) or the Pearson correlation coefficient \(\rho\) which could be calculated using the associated sample estimator \emph{r} are commonly used to compare heterogeneous outcome variables\footnote{\color{red} When using bounded effect sizes such as the Pearson correlation, Fisher Z-transformation is commonly used (Borenstein et al., 2009, pp. 41--43) \color{black}}. If all studies used the same raw measure, such as reaction times, it is possible to directly meta-analyze the studies without standardizing e.g.~using the unstandardized mean difference (UMD, Borenstein et al., 2009, pp. 21--24).

@@

\color{red}

The core of a meta-analysis is combining the results of multiple studies giving more weight to studies that provide a more precise effect estimation. We can find essentially three meta-analysis models: the equal-effects, the fixed-effects, and the random-effects model (Hedges \& Vevea, 1998; Laird \& Mosteller, 1990). The \emph{equal-effects} model assumes that each study included in the meta-analysis is a more or less precise estimation of the true underlying effect (\(\theta\)). In other terms, we are assuming that there is no variability (i.e., heterogeneity) among true effect sizes. On the other side, suppose some study-level characteristics (e.g., participants' age, sex, or socioeconomic status) or the experimental paradigm (e.g., type of memory task or difficulty) could impact the treatment effect. In this case, we have true variability (i.e., heterogeneity) among studies. The \emph{random-effects} model assumes a distribution of real effects with mean \(\mu_{\theta}\) and variance \(\tau^{2}\) thus estimating the heterogeneity. Finally, the \emph{fixed-effects} model estimate the average effect of the included pool of studies ignoring the presence of heterogeneity. While the distinction between the equal and fixed-effects model is theoretical, the estimated model is the exactly the same\footnote{See \href{https://wviechtb.github.io/metafor/reference/misc-models.html}{wviechtb.github.io/metafor/reference/misc-models.html} for a detailed explanation}. The \emph{random-effects} model assumes and estimates heterogeneity among effect sizes leading to different model parameters. From an inferential point of view, the \emph{random-effects} model provides unconditional inference to the population of effect sizes while the \emph{fixed-effects} model estimates the average effect of the selected studies and not population-level parameters. The \emph{equal-effects} model assumes a single true population effect to be estimated. Figure \ref{fig:img-equal-vs-random} depicts the theoretical distinction between the equal and random-effects model. In the presence of heterogeneity, we can include moderators (e.g., the type of experimental setup) into a meta-regression model to explain the effect size heterogeneity. \color{red} An interesting proposal to combine evidence from these different models into a single analysis is called Bayesian model-averaged meta-analysis (Berkhout, Haaf, Gronau, Heck, \& Wagenmakers, 2023; Gronau, Heck, Berkhout, Haaf, \& Wagenmakers, 2021). However, this model is beyond the scope of the tutorial. \color{black}

\color{black}

@@

\caption{\color{red} The difference between the assumptions of the \emph{equal} and \emph{random-effects model}. Each distribution depicts the sampling distribution of \(k = 4\) hypothetical studies (\(i = 1\), \(2\), \ldots, \(4\)) with a certain observed effect size \(y_{i}\) (red squares), sampling variability \(\sigma_{\epsilon_i}^{2}\) and the 95\% confidence interval (black segment). The \emph{equal-effects} plot on the left suggests that each observed effect size has the same underlying true effect \(\theta\) (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be close to the real effect (\(\theta\)). The \emph{random-effects model} on the right suggests that beyond the error term (\(\epsilon_{i}\)), each real effect size is composed of a fixed part (now \(\mu_{\theta}\)) and a random part (\(\delta_{i}\)) sampled from a normal distribution with mean zero and variance \(\tau^{2}\). When \(\tau^{2}\) is zero, the random-effects reduces to a equal-effects model. \color{black}}\label{fig:img-equal-vs-random}
\end{figure}

@@

\color{red}

This tutorial uses several R packages for the simulations, meta-analysis fitting, and figures/tables. For the data manipulation, we used the \emph{tidyverse} (Wickham, 2023) package. For the models fitting we used the \emph{metafor} (Viechtbauer, 2010) package. For figures and tables the \emph{ggplot2} (Wickham et al., 2023) or the \texttt{metafor::forest()} function, \emph{kableExtra} (Zhu, 2021) and \emph{papaja} (Aust \& Barth, 2022) packages. We set the \texttt{seed} for reproducibility of the simulation environment.
\color{black}

@@

\color{red}

Before diving into the specific simulations, in this section, we define the common aspects of all simulations in the following sections. The current paper focuses on the \emph{two-level}, \emph{equal,} and \emph{random-effects} models. All the examples refer to primary studies that assess the efficacy of a treatment by comparing a control and an experimental group. In simulation studies where the purpose is not evaluating effect size estimators, it is convenient to simulate unstandardized effect size measures (see Viechtbauer, 2005, 2007). The estimator is unbiased thus not requiring small sample correction (Hedges, 1981; e.g., Hedges, 1989). Furthermore, the effect size and the sampling variance are independent. Similar to the simulation approach by Viechtbauer (2005, 2007) the \emph{experimental} group (\(T\)) and \emph{control} groups \(C\) are sampled from normal distributions respectively \(T_i \sim \mathcal{N}(\Delta, 1)\) and \(C_i \sim \mathcal{N}(0, 1)\) where \(\Delta\) is the unstandardized mean difference (UMD). The UMD and the sampling variance are calculated using Equations \eqref{eq:effsize} and \eqref{eq:effsizevar} where \(y\) is the estimated value of \(\Delta\). \color{black}

\color{red}

@@

\color{red}

The \texttt{es} is the true effect size (\(\mu_{\theta}\) for the random-effects model and \(\theta\) for the equal-effects model), \texttt{nc} and \texttt{nt} are the sample size for the control and experimental group, \texttt{aggregate} controls if returning the effect size and the corresponding sampling variance or the participant-level data. We can generate a single study with the desired parameters with this function. A suggestion for each simulation step is to generate a large \(n\) to reduce the sampling error and check the recovery of simulated parameters. This is a general strategy that can be applied to every simulation. The \texttt{sim\_studies()} function will iterate through variables in the \texttt{...} argument creating the meta-analysis data frame. The \texttt{mapply()} function is clearly explained in the supplementary materials. \color{black}

@@

Using the \texttt{sim\_study()} function multiple times (with the appropriate adjustments) we can generate a series of studies simulating a data set for a meta-analysis (using the \texttt{sim\_studies()} function). After each example, we will compute the appropriate model (e.g., \emph{equal} or \emph{random-effects}) using the \texttt{metafor} package (Viechtbauer, 2010) to check the recovery of simulated parameters\footnote{\color{red} By default \texttt{metafor} use the Wald \emph{z} test on model parameters but other inference methods are available e.g.~the Knapp and Hartung (2003) method by setting \texttt{test="knha"} within the \texttt{rma()} function. This method is useful especially when the number of studies is low (see \href{https://wviechtb.github.io/metafor/reference/misc-recs.html}{wviechtb.github.io/metafor/reference/misc-recs.html} for more details) \color{black}}. Table \ref{tab:notation-tab} \textcolor{red}{summarizes the notation used in simulations and code in equations and code.}

@@

The most basic model to simulate is the \emph{equal-effects model} (EE). As reported in the previous sections, the \emph{equal-effects model} assumes the presence of a single true effect (\(\theta\)) and the \color{red} observed variability is caused by each effect size being a more or less imprecise estimation of the true effect. \color{black} In other words, the only source of variability is the sampling variability that depends on the variance of the primary studies (i.e., the sample size). Equations \eqref{eq:equal-effects-model1} and \eqref{eq:equal-effects-model2} \textcolor{red}{formalize} the \emph{equal-effects} model.

@@

Each observed effect size \color{red} (\(y_{i}\)) \color{black} is composed of the real effect size \color{red} \(\theta\) \color{black} plus an error term (\(\epsilon_{i}\)) that is sampled from a normal distribution with \(\mu = 0\) and \color{red} \(\sigma^{2} = \sigma_{\epsilon_i}^{2}\) \color{black} (i.e., the known sampling variance of the study \(i\)). As demonstrated in Equation \eqref{eq:effsize}, the increase in sample size will decrease the sampling variability. \textcolor{red}{A study with a extremely large sample size} will essentially have \(\theta\) as the observed effect size. Since we are sampling participants' level data, the error component is already included in the \texttt{sim\_study()} function. To simulate this model in R we can just call the \texttt{sim\_study()} multiple times according to the number of desired studies (\(k\)). In addition, we simulate that each primary study will have a sample size of \(n = 20\) for both groups. We will discuss later the appropriateness of this assumption.

@@

\color{red}

We are using the \(\theta\) parameter to generate random data using the \texttt{sim\_study()} function that will introduce the random error component \(\epsilon_{i}\). Then we can fit the \emph{equal-effects} meta-analysis model with the \texttt{rma} function and \texttt{method\ =\ "EE"} of the \texttt{metafor} package. The parameter we are estimating is \(\theta\) which is close to our simulation value (see Table \ref{tab:res-equal-effects}). Increasing the number of studies (\(k\)) and/or the number of participants in each study (\(n\)) will improve estimation reducing the standard error.
\color{black}

@@

\caption{\label{tab:res-equal-effects}\color{red} Summary of the simulated equal-effects model. The only estimated parameter is the average effect \(\theta\) (\(\beta\)) with the standard error, 95\% confidence interval, and the Wald z-test. The z-test evaluates the null hypothesis that the real effect equals zero.\color{black}}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{ccccc}
\toprule
 & $\beta$ & 95\% CI & z & p\\
\midrule
overall & 0.347 (SE = 0.056) & $[0.237, 0.456]$ & 6.223 & < 0.001\\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{} $k = 30$}\\
\end{tabular}
\end{table}

@@

\caption{\color{red} Forest plots of two simulated \emph{equal-effects models}. On the left, the simulated model has \(nt,nc = 30\) for each included study while on the right the sample size for each study is \(nt,nc = 500\). Given that the data were generated under a \emph{equal-effects model}, when the sample size is high (on the right) each study is aligned on the real effect size because the error component (\(\epsilon_{i}\)) \textcolor{red}{is close to zero}. The average effect size is similar (depending on the random numbers generation) between the two scenarios while the estimation precision (the width of the black diamond) is narrower on the right. \color{black}}\label{fig:equal-effects-high-low-precision}
\end{figure}

@@

Then we can fit the \emph{random-effects} meta-analysis model with the \texttt{rma} function and \texttt{method\ =\ "REML"}. Table \ref{tab:res-random-effects} depicts the model results\footnote{The \texttt{method\ =\ "REML"} is not the only method to estimate a random-effects model. See the \texttt{rma} documentation (\url{https://wviechtb.github.io/metafor/reference/rma.uni.html\#specifying-the-model}) for an overview of other estimation methods} of the \texttt{metafor} package. We are now estimating two parameters: \(\mu_{\theta}\) and \(\tau^{2}\). As for the \color{red} \emph{equal-effects model} \color{black}, increasing the number of studies and/or the number of participants in each study will improve estimation reducing the standard error.

@@

\caption{\label{tab:res-random-effects}\color{red} Summary of the simulated random-effects model: Compared to the equal-effects model, there are more parameters. The \(\beta\) is the average effect (\(\mu_{\theta}\)) with the standard error, 95\% confidence interval, and the Wald z-test. The \(\tau^{2}\) is the estimated heterogeneity and the \(I^{2}\) (explained in Section \ref{re-model}) represents the percentage of total variability due to between-study heterogeneity. \color{black}}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{ccccc}
\toprule
 & $\beta$ & 95\% CI & z & p\\
\midrule
overall & 0.394 (SE = 0.110) & $[0.179, 0.609]$ & 3.596 & < 0.001\\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{} $k = 30$}\\
\multicolumn{5}{l}{\textsuperscript{} $\tau^2 = 0.262$ ($SE = 0.095$)}\\
\multicolumn{5}{l}{\textsuperscript{} $I^2 = 73.480\%$}\\
\end{tabular}
\end{table}

@@

An important aspect of the \emph{random-effects model} is the interplay between the heterogeneity (\(\tau^{2}\)) and the sampling variability (\(\sigma_{i}^{2}\)). As the number of studies \(k\) increases, the estimation of \(\mu_{\theta}\) and \(\tau^{2}\) becomes more precise \color{red} (Blázquez-Rincón, Sánchez-Meca, Botella, \& Suero, 2023; Rubio-Aparicio et al., 2018) \color{black}. Additionally, as the sample size of each study decreases, each \(\delta_{i}\) will be estimated with higher precision but as long as \(\tau^{2} \neq 0\) there will be variability among effect sizes. In other words, increasing the sample size of each study or the number of studies will not affect the value of \(\tau^{2}\) but only the estimation precision (see Borenstein et al., 2009, Chapter 16, Figure 16.6 for a clear explanation). This can be easily demonstrated using the previous simulation, increasing each study sample size. Figure \ref{fig:random-effects-high-low-precision} depicts the same meta-analysis but with different precision in estimating the study \(y_{i}\). Compared to Figure \ref{fig:equal-effects-high-low-precision}, increasing the sample size of primary studies improves the estimation of each study without reducing the between-studies heterogeneity.

@@

\caption{\color{red} Forest plots of two simulated \emph{random-effects model}s. On the left, the simulated model has \(nt,nc = 30\) for each included study, while on the right, the sample size for each study is \(nt,nc = 500\). Compared to Figure \ref{fig:equal-effects-high-low-precision}, data were generated under a \emph{random-effects model}. The estimated average effect is similar between the two scenarios regarding average effect and precision. However, compared to the \emph{equal-effects} simulation, increasing the sample size of primary studies only affects the precision without reducing the true heterogeneity (i.e., \(\tau^{2} \neq 0\)). \color{black}}\label{fig:random-effects-high-low-precision}
\end{figure}

>>>>

\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

@@

\abstract{%
Meta-analysis is a powerful tool to combine evidence from existing literature. Despite several introductory and advanced materials about organizing, conducting, and reporting a meta-analysis, to our knowledge, there are no introductive materials about simulating the most common meta-analysis models. Data simulation is essential for developing and validating new statistical models and procedures. Furthermore, data simulation is a powerful educational tool for understanding a statistical method. In this tutorial, we show how to simulate equal-effects, random-effects, and meta-regression models and illustrate how to estimate statistical power. In the supplementary materials, we extended these simulations also for multilevel and multivariate models. All materials associated with this article can be accessed on Open Science Framework (\url{https://osf.io/54djn/}).
}

@@

A meta-analysis is an essential tool for combining knowledge from multiple studies quantitatively. Meta-analysis is commonly used together with a systematic review of the literature. The meta-analysis has several advantages. Firstly, it allows combining evidence from multiple studies assigning more weight to studies with lower estimation variability. Then using meta-regression, it is possible to include variables (i.e., moderators) to explain the observed heterogeneity (Borenstein, Hedges, Higgins, \& Rothstein, 2009, pp. 187--203). More recently, location-scale models have been developed to include predictors also on the residual heterogeneity (i.e., Viechtbauer \& López-López, 2022). Finally, considering the replication crisis, there are statistical methods to determine the presence and extent of the publication bias. Despite the advantages, meta-analysis implementation is not always straightforward, especially for complex data structures. Additionally, the fact that there are several introductory and advanced resources to understand meta-analysis (Borenstein et al., 2009; Harrer, Cuijpers, Furukawa, \& Ebert, 2021; Schmid, Stijnen, \& White, 2022), to our knowledge, there are no introductory resources about how to simulate realistic meta-analytic data.

Simulating data has several advantages because it requires understanding the statistical method and the data-generation process. Furthermore, data simulation is the primary tool when it comes to evaluating a new analysis method, estimating the statistical power, or understanding the long-run behavior of our data generation process (Gelman \& Hill, 2006, pp. 155--176; Gelman, Hill, \& Vehtari, 2020, pp. 69--76; Ingalls, 2011). A recent paper by DeBruine and Barr (2021), which deeply inspired the current work, proposed a stimulating way to understand linear mixed-effects models via data simulation. Simulating data is also a powerful educational tool within this framework.

For these reasons, this work aims to introduce the basic concepts of meta-analysis and Monte Carlo simulations for equal-effects, random-effects, and meta-regression models with applications also to statistical power calculation. In the first section, we will introduce basic concepts of the meta-analysis that are useful for setting up the simulation. We will evaluate the effect size, variance calculation, and the equal vs.~random effects model distinction. Then we will describe how to simulate data for these models and simulate a meta-regression with categorical and numerical predictors. Finally, we will introduce the power analysis extending the previous examples to estimate the statistical power. We used the R statistical programming language (version 4.3.1; R Core Team, 2023).

The aim of the tutorial is not providing a complete theoretical introduction to meta-analysis but presenting the core topics using a simulation-based approach. Readers expert in conducting meta-analyses could benefit more from the proposed approach in terms of simulation setup and coding strategies. Readers without experience in meta-analysis could benefit both from the theoretical introduction and the simulation approach. However, for a better and extensive overview of meta-analysis topics the reader could refer to Harrer, Cuijpers, and Ebert (2019), and Borenstein et al. (2009).

We assume the reader is familiar with the basic concepts of R, but core functions will be explained. Code and materials are available on the OSF repository (\url{https://osf.io/54djn/}). A theoretical introduction, simulation examples for multivariate and multilevel models, and more details about the coding approach are available in the supplementary materials.

@@

As an example that will be used throughout the paper, we consider the efficacy of memory training in improving memory performance during a cognitive task. The typical primary study will collect data from a group of participants receiving the memory training (\emph{experimental group}) and another group receiving a control treatment (\emph{control group}). The focus of the meta-analysis is collecting multiple studies with similar aims and methods and estimating the average effect of memory training. Despite differences in the type of cognitive task or experimental setup, each primary study collects an \emph{experimental group} (\(n_{T}\)) and \emph{control group} (\(n_{C}\)) and computes the average performance (\emph{experimental group} \(\overline T\) and \emph{control group} \(\overline C\)) and standard deviations (\emph{experimental group} \(s_{T}\) and \emph{control group}
\(s_{C}\)).

@@

The first step of a meta-analysis is to extract information from included studies. This common measure should give an immediate idea of the direction (i.e., the treatment improves or reduces performance) and the size of the effect. Standardized effect size measures (see Lakens, 2013 for an overview), such as the Standardized Mean Difference (SMD), which could be estimated by Cohen's \(d\) (Cohen, 1988) or the Pearson correlation coefficient \(\rho\) which could be calculated using the associated sample estimator \emph{r} are commonly used to compare heterogeneous outcome variables\footnote{ When using bounded effect sizes such as the Pearson correlation, Fisher Z-transformation is commonly used (Borenstein et al., 2009, pp. 41--43)}. If all studies used the same raw measure, such as reaction times, it is possible to directly meta-analyze the studies without standardizing e.g.~using the unstandardized mean difference (UMD, Borenstein et al., 2009, pp. 21--24).

@@

The core of a meta-analysis is combining the results of multiple studies giving more weight to studies that provide a more precise effect estimation. We can find essentially three meta-analysis models: the equal-effects, the fixed-effects, and the random-effects model (Hedges \& Vevea, 1998; Laird \& Mosteller, 1990). The \emph{equal-effects} model assumes that each study included in the meta-analysis is a more or less precise estimation of the true underlying effect (\(\theta\)). In other terms, we are assuming that there is no variability (i.e., heterogeneity) among true effect sizes. On the other side, suppose some study-level characteristics (e.g., participants' age, sex, or socioeconomic status) or the experimental paradigm (e.g., type of memory task or difficulty) could impact the treatment effect. In this case, we have true variability (i.e., heterogeneity) among studies. The \emph{random-effects} model assumes a distribution of real effects with mean \(\mu_{\theta}\) and variance \(\tau^{2}\) thus estimating the heterogeneity. Finally, the \emph{fixed-effects} model estimate the average effect of the included pool of studies ignoring the presence of heterogeneity. While the distinction between the equal and fixed-effects model is theoretical, the estimated model is the exactly the same\footnote{See \href{https://wviechtb.github.io/metafor/reference/misc-models.html}{wviechtb.github.io/metafor/reference/misc-models.html} for a detailed explanation}. The \emph{random-effects} model assumes and estimates heterogeneity among effect sizes leading to different model parameters. From an inferential point of view, the \emph{random-effects} model provides unconditional inference to the population of effect sizes while the \emph{fixed-effects} model estimates the average effect of the selected studies and not population-level parameters. The \emph{equal-effects} model assumes a single true population effect to be estimated. Figure \ref{fig:img-equal-vs-random} depicts the theoretical distinction between the equal and random-effects model. In the presence of heterogeneity, we can include moderators (e.g., the type of experimental setup) into a meta-regression model to explain the effect size heterogeneity. An interesting proposal to combine evidence from these different models into a single analysis is called Bayesian model-averaged meta-analysis (Berkhout, Haaf, Gronau, Heck, \& Wagenmakers, 2023; Gronau, Heck, Berkhout, Haaf, \& Wagenmakers, 2021). However, this model is beyond the scope of the tutorial.

@@

\caption{The difference between the assumptions of the \emph{equal} and \emph{random-effects model}. Each distribution depicts the sampling distribution of \(k = 4\) hypothetical studies (\(i = 1\), \(2\), \ldots, \(4\)) with a certain observed effect size \(y_{i}\) (pink squares), sampling variability \(\sigma_{\epsilon_i}^{2}\) and the 95\% confidence interval (black segment). The \emph{equal-effects} plot on the left suggests that each observed effect size has the same underlying true effect \(\theta\) (each distribution has the same mean) with a different degree of precision (e.g., Study 1 is more precise than Study 3). In practice, each study has a different observed effect size where studies with high precision (i.e., narrow sampling distributions) will be close to the real effect (\(\theta\)). The \emph{random-effects model} on the right suggests that beyond the error term (\(\epsilon_{i}\)), each real effect size is composed of a fixed part (now \(\mu_{\theta}\)) and a random part (\(\delta_{i}\)) sampled from a normal distribution with mean zero and variance \(\tau^{2}\). When \(\tau^{2}\) is zero, the random-effects reduces to a equal-effects model.}\label{fig:img-equal-vs-random}
\end{figure}

@@

This tutorial uses several R packages for the simulations, meta-analysis fitting, and figures/tables. For the data manipulation, we used the \emph{tidyverse} (Wickham, 2023) package. For the models fitting we used the \emph{metafor} (Viechtbauer, 2010) package. For figures and tables the \emph{ggplot2} (Wickham et al., 2023) or the \texttt{metafor::forest()} function, \emph{kableExtra} (Zhu, 2021) and \emph{papaja} (Aust \& Barth, 2022) packages. We set the \texttt{seed} for reproducibility of the simulation environment.

@@

Before diving into the specific simulations, in this section, we define the common aspects of all simulations in the following sections. The current paper focuses on the \emph{two-level}, \emph{equal,} and \emph{random-effects} models. All the examples refer to primary studies that assess the efficacy of a treatment by comparing a control and an experimental group. In simulation studies where the purpose is not evaluating effect size estimators, it is convenient to simulate unstandardized effect size measures (see Viechtbauer, 2005, 2007). The estimator is unbiased thus not requiring small sample correction (Hedges, 1981; e.g., Hedges, 1989). Furthermore, the effect size and the sampling variance are independent. Similar to the simulation approach by Viechtbauer (2005, 2007) the \emph{experimental} group (\(T\)) and \emph{control} groups \(C\) are sampled from normal distributions respectively \(T_i \sim \mathcal{N}(\Delta, 1)\) and \(C_i \sim \mathcal{N}(0, 1)\) where \(\Delta\) is the unstandardized mean difference (UMD). The UMD and the sampling variance are calculated using Equations \eqref{eq:effsize} and \eqref{eq:effsizevar} where \(y\) is the estimated value of \(\Delta\).

@@

The \texttt{es} is the true effect size (\(\mu_{\theta}\) for the random-effects model and \(\theta\) for the equal-effects model), \texttt{nc} and \texttt{nt} are the sample size for the control and experimental group, \texttt{aggregate} controls if returning the effect size and the corresponding sampling variance or the participant-level data. We can generate a single study with the desired parameters with this function. A suggestion for each simulation step is to generate a large \(n\) to reduce the sampling error and check the recovery of simulated parameters. This is a general strategy that can be applied to every simulation. The \texttt{sim\_studies()} function will iterate through variables in the \texttt{...} argument creating the meta-analysis data frame. The \texttt{mapply()} function is clearly explained in the supplementary materials.

@@

Using the \texttt{sim\_study()} function multiple times (with the appropriate adjustments) we can generate a series of studies simulating a data set for a meta-analysis (using the \texttt{sim\_studies()} function). After each example, we will compute the appropriate model (e.g., \emph{equal} or \emph{random-effects}) using the \texttt{metafor} package (Viechtbauer, 2010) to check the recovery of simulated parameters\footnote{ By default \texttt{metafor} use the Wald \emph{z} test on model parameters but other inference methods are available e.g.~the Knapp and Hartung (2003) method by setting \texttt{test="knha"} within the \texttt{rma()} function. This method is useful especially when the number of studies is low (see \href{https://wviechtb.github.io/metafor/reference/misc-recs.html}{wviechtb.github.io/metafor/reference/misc-recs.html} for more details)}. Table \ref{tab:notation-tab} summarizes the notation used in simulations and code in equations and code.

@@

The most basic model to simulate is the \emph{equal-effects model} (EE). As reported in the previous sections, the \emph{equal-effects model} assumes the presence of a single true effect (\(\theta\)) and the observed variability is caused by each effect size being a more or less imprecise estimation of the true effect. In other words, the only source of variability is the sampling variability that depends on the variance of the primary studies (i.e., the sample size). Equations \eqref{eq:equal-effects-model1} and \eqref{eq:equal-effects-model2} formalize the \emph{equal-effects} model.

@@

Each observed effect size (\(y_{i}\)) is composed of the real effect size \(\theta\) plus an error term (\(\epsilon_{i}\)) that is sampled from a normal distribution with \(\mu = 0\) and \(\sigma^{2} = \sigma_{\epsilon_i}^{2}\) (i.e., the known sampling variance of the study \(i\)). As demonstrated in Equation \eqref{eq:effsize}, the increase in sample size will decrease the sampling variability. A study with a extremely large sample size will essentially have \(\theta\) as the observed effect size. Since we are sampling participants' level data, the error component is already included in the \texttt{sim\_study()} function. To simulate this model in R we can just call the \texttt{sim\_study()} multiple times according to the number of desired studies (\(k\)). In addition, we simulate that each primary study will have a sample size of \(n = 20\) for both groups. We will discuss later the appropriateness of this assumption.

@@

We are using the \(\theta\) parameter to generate random data using the \texttt{sim\_study()} function that will introduce the random error component \(\epsilon_{i}\). Then we can fit the \emph{equal-effects} meta-analysis model with the \texttt{rma} function and \texttt{method\ =\ "EE"} of the \texttt{metafor} package. The parameter we are estimating is \(\theta\) which is close to our simulation value (see Table \ref{tab:res-equal-effects}). Increasing the number of studies (\(k\)) and/or the number of participants in each study (\(n\)) will improve estimation reducing the standard error.

@@

\caption{\label{tab:res-equal-effects}Summary of the simulated equal-effects model. The only estimated parameter is the average effect \(\theta\) (\(\beta\)) with the standard error, 95\% confidence interval, and the Wald z-test. The z-test evaluates the null hypothesis that the real effect equals zero.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{ccccc}
\toprule
 & $\beta$ & 95\% CI & z & p\\
\midrule
overall & 0.347 (SE = 0.056) & $[0.237, 0.456]$ & 6.223 & < 0.001\\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{} $k = 30$}\\
\end{tabular}
\end{table}

@@

\caption{Forest plots of two simulated \emph{equal-effects models}. On the left, the simulated model has \(nt,nc = 30\) for each included study while on the right the sample size for each study is \(nt,nc = 500\). Given that the data were generated under a \emph{equal-effects model}, when the sample size is high (on the right) each study is aligned on the real effect size because the error component (\(\epsilon_{i}\)) is close to zero. The average effect size is similar (depending on the random numbers generation) between the two scenarios while the estimation precision (the width of the black diamond) is narrower on the right.}\label{fig:equal-effects-high-low-precision}
\end{figure}

@@

Then we can fit the \emph{random-effects} meta-analysis model with the \texttt{rma} function and \texttt{method\ =\ "REML"}. Table \ref{tab:res-random-effects} depicts the model results\footnote{The \texttt{method\ =\ "REML"} is not the only method to estimate a random-effects model. See the \texttt{rma} documentation (\url{https://wviechtb.github.io/metafor/reference/rma.uni.html\#specifying-the-model}) for an overview of other estimation methods} of the \texttt{metafor} package. We are now estimating two parameters: \(\mu_{\theta}\) and \(\tau^{2}\). As for the \emph{equal-effects model} , increasing the number of studies and/or the number of participants in each study will improve estimation reducing the standard error.

@@

\caption{\label{tab:res-random-effects}Summary of the simulated random-effects model: Compared to the equal-effects model, there are more parameters. The \(\beta\) is the average effect (\(\mu_{\theta}\)) with the standard error, 95\% confidence interval, and the Wald z-test. The \(\tau^{2}\) is the estimated heterogeneity and the \(I^{2}\) (explained in Section \ref{re-model}) represents the percentage of total variability due to between-study heterogeneity.}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{ccccc}
\toprule
 & $\beta$ & 95\% CI & z & p\\
\midrule
overall & 0.394 (SE = 0.110) & $[0.179, 0.609]$ & 3.596 & < 0.001\\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{} $k = 30$}\\
\multicolumn{5}{l}{\textsuperscript{} $\tau^2 = 0.262$ ($SE = 0.095$)}\\
\multicolumn{5}{l}{\textsuperscript{} $I^2 = 73.480\%$}\\
\end{tabular}
\end{table}

@@

An important aspect of the \emph{random-effects model} is the interplay between the heterogeneity (\(\tau^{2}\)) and the sampling variability (\(\sigma_{i}^{2}\)). As the number of studies \(k\) increases, the estimation of \(\mu_{\theta}\) and \(\tau^{2}\) becomes more precise (Blázquez-Rincón, Sánchez-Meca, Botella, \& Suero, 2023; Rubio-Aparicio et al., 2018) . Additionally, as the sample size of each study decreases, each \(\delta_{i}\) will be estimated with higher precision but as long as \(\tau^{2} \neq 0\) there will be variability among effect sizes. In other words, increasing the sample size of each study or the number of studies will not affect the value of \(\tau^{2}\) but only the estimation precision (see Borenstein et al., 2009, Chapter 16, Figure 16.6 for a clear explanation). This can be easily demonstrated using the previous simulation, increasing each study sample size. Figure \ref{fig:random-effects-high-low-precision} depicts the same meta-analysis but with different precision in estimating the study \(y_{i}\). Compared to Figure \ref{fig:equal-effects-high-low-precision}, increasing the sample size of primary studies improves the estimation of each study without reducing the between-studies heterogeneity.

@@

\caption{Forest plots of two simulated \emph{random-effects model}s. On the left, the simulated model has \(nt,nc = 30\) for each included study, while on the right, the sample size for each study is \(nt,nc = 500\). Compared to Figure \ref{fig:equal-effects-high-low-precision}, data were generated under a \emph{random-effects model}. The estimated average effect is similar between the two scenarios regarding average effect and precision. However, compared to the \emph{equal-effects} simulation, increasing the sample size of primary studies only affects the precision without reducing the true heterogeneity (i.e., \(\tau^{2} \neq 0\)).}\label{fig:random-effects-high-low-precision}
\end{figure}
